---
title: "Chapter 4: ANOVAs Part III"
author: "Andy Supple"
format:
  html:
    code-fold: false
editor: source
execute:
  warning: false
  message: false
references: references.bib
---

# Repeated-Measures ANOVA

Up to this point, the ANOVAs we have examined have focused primarily on **between-group differences** in an outcome. These analyses compare mean scores across different groups of participants and are appropriate for between-subjects designs.

Another important member of the ANOVA family is the **repeated-measures ANOVA** (RM ANOVA).

A repeated-measures design involves collecting **multiple assessments from the same participant**. Rather than comparing mean scores between different groups of people, repeated-measures ANOVA evaluates whether mean scores differ **within the same participants across multiple conditions or time points**.

Examples from health and human sciences research include:

- Do participants report different levels of exertion when running on a treadmill while talking with a friend, listening to music, listening to an audiobook, or listening to nothing?
- Do participants beginning a new weight-training regimen report higher levels of self-confidence at three and six months compared with baseline?
- Do parents who attend a divorce education program show reductions in interparental conflict from the beginning to the end of a 10-week intervention?
- Do participants report increased fruit and vegetable consumption after attending a nutrition education program compared with their consumption before the program?

In each example, the primary research question concerns whether scores change within individuals across conditions or time points. Repeated measures are commonly used to evaluate change following an intervention, but they can also be used whenever researchers wish to compare participants' ratings, preferences, or performance across multiple conditions.

## Mixed Designs

Many studies combine repeated measures with a traditional between-subjects grouping variable. These designs are called **mixed designs** because they include both:

- A **within-subjects factor**, representing repeated observations from the same participants.
- A **between-subjects factor**, representing differences between groups of participants.

Researchers frequently use mixed designs to evaluate intervention effectiveness. In these studies:

- One group receives a treatment or intervention that is expected to produce change.
- A comparison or control group does not receive the intervention.
- Participants in both groups are measured repeatedly over time.

Mixed designs allow researchers to address three questions:

1. **Within-subjects effect:** Do participants change over time?
2. **Between-subjects effect:** Do the groups differ overall?
3. **Interaction effect:** Does the pattern of change over time differ across groups?

For example, consider a study examining a six-week strength-training program:

- Does maximum leg-press strength increase from baseline to six weeks?
- Do participants in the training program have greater strength overall than participants in a comparison group?
- Does the increase in strength over time differ between the training and comparison groups?

Researchers might expect the two groups to be similar at baseline but differ after six weeks. Statistically, this prediction corresponds to an **interaction between time and group membership**, indicating that the amount of change over time depends on whether participants participated in the strength-training program.

# Connection to the Paired-Samples *t* Test

The paired-samples *t* test evaluates whether the mean difference between two measurements equals zero. Rather than comparing two independent groups, the same participant contributes two observations.

Common examples include:

- Pretest versus posttest
- Before treatment versus after treatment
- Beginning of semester versus end of semester
- Parent report versus child report
- Two conditions measured on the same person

For an independent-samples *t* test:

$$
\text{Person A is unrelated to Person B}
$$

For a paired-samples *t* test:

$$
\text{Observation 1 and Observation 2 belong to the same person}
$$

The observations are linked, or paired. Ignoring the pairing discards information and generally reduces statistical power.

## Difference Scores

The paired-samples *t* test focuses on the difference score:

$$
D = X_1 - X_2
$$

where:

- $X_1$ is the first measurement.
- $X_2$ is the second measurement.
- $D$ is the difference score.

These could represent

- Differences in participant scores before and after some experience
- Differences in wives' and husbands' report of marital conflict
- Differences in a participant's motivation to excercise versus play video games

## Null and Alternative Hypotheses

The null hypothesis is:

$$
H_0: \mu_D = 0
$$

The alternative hypothesis is:

$$
H_A: \mu_D \neq 0
$$

where $\mu_D$ is the population mean difference.

## Test Statistic

The paired-samples *t* statistic is:

$$
t =
\frac{\bar{D}}
{s_D / \sqrt{n}}
$$

where:

- $\bar{D}$ is the mean difference score.
- $s_D$ is the standard deviation of the difference scores.
- $n$ is the number of pairs.

The formula can be interpreted as:

$$
t =
\frac{\text{Signal}}
{\text{Noise}}
$$

The signal is the average change. The noise is the uncertainty in that change.

## Degrees of Freedom

For a paired-samples *t* test:

$$
df = n - 1
$$

For example, if there are 30 participants:

$$
df = 29
$$

Degrees of freedom depend on the number of pairs, not the total number of observations.

## Assumptions

### 1. Paired Observations

Scores must be meaningfully linked.

Examples include:

- The same person measured twice
- Married couples
- Parent-child dyads

### 2. Independence Between Pairs

One participant's scores should not influence another participant's scores.

### 3. Approximate Normality of Difference Scores

The normality assumption applies to:

$$
D = X_1 - X_2
$$

The original variables do not each need to be perfectly normally distributed.

## Why Is Pairing Helpful?

Suppose people differ substantially in baseline characteristics. Comparing participants with themselves removes much of this person-to-person variability.

As a result:

- Standard errors may decrease.
- Statistical power may increase.
- Smaller effects may become easier to detect.

This is one reason repeated-measures designs are often efficient.

## Effect Size

A common effect size for paired data is:

$$
d =
\frac{\bar{D}}{s_D}
$$

General guidelines are:

| $d$ | Interpretation |
|---:|:---|
| 0.20 | Small |
| 0.50 | Medium |
| 0.80 | Large |

Effect size should always be interpreted separately from statistical significance.

## R Code
```
t.test(
  data$posttest,
  data$pretest,
  paired = TRUE)
  
```
# Repeated-Measures ANOVA

Repeated-measures ANOVA extends the paired-samples *t* test to situations involving three or more repeated conditions or time points. A two-level repeated-measures ANOVA is mathematically equivalent to a paired-samples *t* test.

A mixed-design ANOVA additionally includes one or more between-subjects factors.

## Example 1: One-Factor Repeated-Measures ANOVA

### Research Scenario

Suppose exercise researchers are interested in whether different auditory distractions influence perceived exertion during exercise.

Participants complete a 20-minute treadmill run under four conditions:

- Talking with a friend
- Listening to music
- Listening to an audiobook
- No audio stimulus

After each run, participants rate their perceived exertion on a scale ranging from 1 to 20.

### Why Use a Repeated-Measures ANOVA?

A traditional one-way ANOVA would not be appropriate because the observations are not independent. Each participant contributes four observations, one for each listening condition.

A repeated-measures ANOVA is appropriate because:

- The same participants experience all four conditions.
- Each participant serves as their own control.
- Stable individual differences can be separated from the within-person error term.
- The analysis evaluates whether mean exertion differs across listening conditions.

The within-subjects factor is `condition`, with four levels:

- Talking
- Music
- Audiobook
- Silence

The outcome variable is perceived exertion.

### Simulate Data

The simulation includes a participant-level effect so that scores from the same participant are positively correlated across conditions.

```{r}
library(tidyverse)

set.seed(627)

n <- 30

participant_effect <- rnorm(n, mean = 0, sd = 1)

dat <- data.frame(
  id = factor(1:n),
  talking = 14.5 + participant_effect + rnorm(n, 0, 1.5),
  music = 13.0 + participant_effect + rnorm(n, 0, 1.5),
  audiobook = 14.0 + participant_effect + rnorm(n, 0, 1.5),
  silence = 15.5 + participant_effect + rnorm(n, 0, 1.5))

head(dat)

```

::: {.callout-note}

Note that these are *wide* format data. There is one row per id # and then there are columns for each of the conditions, the rows contain the single id # and then scores on perceived exertion across the conditions. For our analysis, these data need to be "turned" or reformatted into a person-period or long dataframe structure.

:::

### Reshape to Long Format using pivot_longer 

```{r}
long_dat <- dat |>
  pivot_longer(cols = talking:silence, 
    names_to = "condition",
    values_to = "exertion") |>
  mutate(condition = factor(condition,levels = c("talking", "music", 
                                            "audiobook", "silence")))

# pivot_longer function takes the columns of talking:silence or talking, music, 
# audiobook, silence and makes them longer, moves their names to a column called
# 'condition' and then takes the values in those original variables and puts 
# them in a new column called 'exertion'
# mutate is a modification to avariable, in this care creatina a factor variable
# out of 'condition' and given it levels

head(long_dat)
```

::: {.callout-note title = "Turning data details"}

In the original wide-format data, each row represents one participant and each condition is represented by a separate column. In the long-format data, each participant has one row for each repeated condition.

The above R code uses *pivot_longer* which is a tidyr function. It takes data in a wide-format and "pivots" it to a long format.

In the long format, the participant id is repeated - in this case four times - and each participant has a level in "condition" because each participant is "exposed" to the same condition. Then there are four exertion scores for each participant and the RM-ANOVA will focus on if there are mean differences across conditions but **within** participants. 

:::

### Examine Means by Condition

```{r}

# This uses dplyr functions
# Group by = put together scores in a grouping, in this case condition across 
# participants
# summarize across n = n (uses the n from the dataframe), creates a new variable 
# called mean_exertion, computes that via the mean of the variable (exertion)
# same for sd = standard deviation
# groups = "drop" just stops the grouping to avoid any dplyr warnings

long_dat |>
  group_by(condition) |>
  summarise(n = n(),mean_exertion = mean(exertion),
    sd_exertion = sd(exertion),
    .groups = "drop")
```
Looking at these means, appears as if more exertion reported when exercising in silence, lowest with music

### Conduct the Repeated-Measures ANOVA

```{r}
model_exer <- aov(
  exertion ~ condition + Error(id / condition),
  data = long_dat
)

summary(model_exer)
```
These results suggest that the F value is statistically significant, meaning that there is at least one mean comparison across the conditions that is different from another.

::: {.callout-note title="What does Error(id / condition) mean?"}

Repeated-measures data contain observations that are related because they come from the same participant. We therefore need to tell R that the four condition scores for a given participant should not be treated as independent observations.

The expression Error(id / condition) specifies the error structure for the repeated-measures ANOVA. It tells aov() to account for variability associated with participants and with participant-by-condition differences when testing the effect of condition.

:::

### Follow-Up Pairwise Comparisons

Just like with One-Way ANOVAs, when there is a statistically significant F test, follow-up comparisons are helpful to determine which specific conditions varied in mean levels of perceived exertion.

```{r}
library(emmeans)

emmeans(
  model_exer,
  pairwise ~ condition)
```
*Interpretation*: In terms of overall means, perceived exertion was highest in the **silence** group, followed by the talking group, then the audiobook, and then the music group (see emmeans above). There were statistically significant mean differences across conditions with perceived exertion being higher in the silence condition when compared to the other three groups. For example, the result ```music - silence       -2.851 0.374 87  -7.614 <0.0001``` suggests that - on average - participants reported exertion to be 2.851 points lower in the music condition then when in the silence condition. That estimate has a standard error of .374, there are 87 degrees of freedom for this specific *t* test, and the critical value has a $p-value$ of < .0001.

## Example 2: Mixed-Design ANOVA

### Research Scenario

Researchers develop a six-week strength-training intervention and wish to determine whether participation in the program is associated with greater improvement in leg-press strength than participation in a comparison group. *Untrained* athletes who currently do not lift weights were first recruited and then randomly assigned to one of two conditions: the strength training program and a control group. The experimental group was led by a personal trainer and coached on how to properly work to muscle failure with proper form. The control group was asked to work on their leg press strength by going to the recreation center at a university. Participants in both conditions were rewarded with $10 each time they "swiped in" at the student recreation center using their university provided ID.

Participants are assessed at:

- Baseline
- Six weeks

Participants belong to one of two groups:

- Training program
- Control group

### Design

The repeated-measures factor is `time`, with two levels:

- Baseline
- Six weeks

The between-subjects factor is `group`, with two levels:

- Training
- Control

A mixed-design ANOVA addresses three questions:

1. Do participants change over time?
2. Do groups differ overall?
3. Does the amount of change over time differ between groups?

The third question is usually the primary intervention question because it evaluates whether improvement differs between the training and control groups.

### Hypothesized Means

| Group | Baseline | Six Weeks |
|:---|---:|---:|
| Control | 100 | 104 |
| Training | 100 | 125 |

Both groups are expected to begin at approximately the same strength level, but the training group is expected to improve more over time.

The simulation includes a participant-level random effect that is added to both baseline and posttest scores. Because the same effect contributes to both measurements, scores from the same participant are correlated.

```{r, echo=F}
set.seed(627)

n_group <- 25

control_id <- 1:n_group
training_id <- (n_group + 1):(2 * n_group)

control_subject_effect <- rnorm(n_group, 0, 8)
training_subject_effect <- rnorm(n_group, 0, 8)

control <- data.frame(
  id = control_id,
  group = "Control",
  pre = 100 + control_subject_effect + rnorm(n_group, 0, 5),
  post = 104 + control_subject_effect + rnorm(n_group, 0, 5)
)

training <- data.frame(
  id = training_id,
  group = "Training",
  pre = 100 + training_subject_effect + rnorm(n_group, 0, 5),
  post = 125 + training_subject_effect + rnorm(n_group, 0, 5)
)

training_df <- bind_rows(control, training) |>
  mutate(
    group = factor(group, levels = c("Control", "Training"))
  )

head(training_df)
```

### Next step is reshaping or turning the data 

```{r}
long_training_df <- training_df |>
  pivot_longer( cols = pre:post,
    names_to = "time",
    values_to = "strength") |>
  mutate(
    time = factor(
      time,
      levels = c("pre", "post"),
      labels = c("Baseline", "Six Weeks")))

head(long_training_df)
```

This *tibble* shows repeated id numbers in the first column, and then baseline and six weeks strength scores for the first 3 participants in the control group.

### Examine Cell Means

```{r}
long_training_df |>
  group_by(group, time) |>
  summarise(
    n = n(),
    mean_strength = mean(strength),
    sd_strength = sd(strength),
    .groups = "drop"
  )
```
These results suggest that the difference in mean strength from baseline to six weeks looks small in the control group and looks much bigger in the training group. That would imply a *time X condition* interaction where the change from baseline to six weeks depends on group.

### Conduct the Mixed-Design ANOVA using the `afex` package

::: {.callout-note title="The 'afex' package for RM-ANOVAs"}

afex::aov_ez() is designed for repeated-measures and mixed-design ANOVA. By default, it uses Type III sums of squares and sum-to-zero coding for categorical predictors. This is a common choice for factorial and mixed designs, especially when the design is not perfectly balanced. In balanced designs, Type II and Type III tests are often similar; in unbalanced designs, the choice of SS type can affect the results, so it should be reported clearly.

`afex` is also useful because it can:

- Report appropriate corrections for violations of sphericity. (*see below*)
- Provide Greenhouse-Geisser and Huynh-Feldt adjusted results when needed. (*see below*)
- Report effect sizes such as generalized eta squared.
- Handle complex repeated-measures error structures more transparently than a manually specified `aov()` model.
- Work well with `emmeans()` for follow-up comparisons.

:::

```{r}
library(afex)

model_strength <- aov_ez(
  id = "id",
  dv = "strength",
  within = "time",                   # Time varies within participants
  between = "group",                 # group varies between participants only
  data = long_training_df
)

anova(model_strength)
```

In a scenario like this, where these is a statistically significant interaction, the main effects should be interpreted with caution because those are misleading. 

In order to understand better the nature of the interaction, it is often helpful to create a plot of conditional means in the outcome (strength) across the different levels of the two factors, time and group. 


### Visualize the Interaction

```{r}
long_training_df |>
  group_by(group, time) |>
  summarise(
    mean_strength = mean(strength),
    .groups = "drop"
  ) |>
  ggplot(
    aes(
      x = time,
      y = mean_strength,
      color = group,
      group = group
    )
  ) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    x = "Time",
    y = "Mean Leg-Press Strength",
    color = "Group"
  ) +
  theme_minimal()
```
This visual makes it clear that the time **effect** in the control group is very small, whereas it is much larger in the training group (blue line). This figure plots out conditional means, or means at different levels of two factors

### Interpreting the Three Effects

Main Effect of Time: whether participants change from baseline to six weeks, averaging across groups.

Main Effect of Group: whether the training and control groups differ overall, averaging across time.

Time-by-Group Interaction: whether the amount of change over time depends on group membership.

In an intervention study, this interaction is often the primary research question. A significant interaction indicates that the groups have different patterns of change. The direction and practical meaning of that interaction should be examined using cell means, a graph, and follow-up comparisons.

In the case of a statistically significant interaction, beyond a figure, we can also look at mean comparisons using emmeans and compare groups at each time point with this:


```{r}
emmeans(
  model_strength,
  pairwise ~ group | time)   # the vertical bar | means nested in
```
Which shows us that *at baseline* there was *no* different in strength across groups.

At six weeks, however, as hypothesized, there was a group difference favoring the training group.

Researchers can also compare time points within each group:

```{r}
emmeans(
  model_strength,
  pairwise ~ time | group)
```
And with these results we can see that within the training group there is a large difference score from baseline to six weeks, and a small difference score from baseline to six weeks in the control group.

Together these post hoc results and the figure give us a nice conceptual sense of how effective the intervention program was. 

# Repeated-Measures ANOVA with More Than Two Time Points

Many studies include more than two measurement occasions. These studies may involve interventions related to health, nutrition, exercise, parenting, or education. They may also examine developmental change and whether that change differs across groups.

Traditional repeated-measures ANOVA can be useful for simple, balanced designs. However, many contemporary researchers use linear mixed models for more complex longitudinal data. Linear mixed models can:

- Accommodate some incomplete data under assumptions about the missing-data mechanism.
- Model participant-level random effects.
- Represent more flexible covariance structures.
- Avoid requiring sphericity for the repeated-measures outcome.

Mixed models do not automatically solve missing-data problems. Their validity depends on model assumptions, including assumptions about how data are missing.

Mixed models (*later in semester*) overcome two other significant flaws with RM ANOVAs.

### Missing Data

Traditional repeated-measures ANOVA generally requires complete data for every participant across all measurement occasions. If a participant is missing even a single observation, many implementations remove that participant from the analysis.

For example, suppose a participant completes assessments at baseline and three months but misses the six-month assessment. In a traditional complete-case repeated-measures ANOVA, that participant would typically be excluded.

This approach, known as listwise deletion, can create several problems:

- Reduced sample size and statistical power.
- Potential bias if missingness is related to the outcome or predictors.
- Loss of information from participants who completed most of the study.

Linear mixed models can often incorporate participants with incomplete data, provided the model and missing-data assumptions are reasonable.

### Sphericity

Sphericity is an assumption concerning the variances of difference scores among the levels of a within-subjects factor.

For a study with three measurement occasions:

- Baseline
- Three months
- Six months

the relevant difference scores include:

- Baseline minus three months
- Baseline minus six months
- Three months minus six months

Sphericity requires the variances of these difference scores to be approximately equal.

### Why Does Sphericity Matter?

Violations of sphericity can inflate the Type I error rate. This means that researchers may be more likely to conclude that a statistically significant effect exists when the effect is not actually present in the population.

Sphericity is relevant for within-subjects effects with three or more levels. With only two levels, there is only one set of difference scores, so sphericity is automatically satisfied.

| Design | Sphericity Required? |
|:---|:---:|
| Pretest versus posttest | No |
| Baseline, three months, six months | Yes |
| Four listening conditions | Yes |

### Testing Sphericity

The most common test of sphericity is Mauchly's test.

The hypotheses are:

$$
H_0: \text{Sphericity is satisfied}
$$

$$
H_A: \text{Sphericity is violated}
$$

Interpretation is usually described as follows:

- A nonsignificant result, $p > .05$, provides no evidence against sphericity. It does not prove that the assumption is exactly true.
- A significant result, $p < .05$, indicates evidence that sphericity has been violated.

Because Mauchly's test can be sensitive to sample size, it should be considered along with the design, the data, and the corrected results.

### Adjusting for Violations of Sphericity

When sphericity is violated, researchers typically apply a correction to the degrees of freedom used in the *F* test.

#### Greenhouse-Geisser Correction

The Greenhouse-Geisser correction is commonly reported. It:

- Reduces the degrees of freedom.
- Produces a more conservative test.
- Usually produces a larger *p* value.
- Provides protection against inflated Type I error.

#### Huynh-Feldt Correction

The Huynh-Feldt correction is less conservative than the Greenhouse-Geisser correction. It adjusts the degrees of freedom to a lesser extent and is sometimes used when the violation is relatively mild.

### Reporting Corrected Results

When Mauchly's test indicates that sphericity has been violated, researchers should report the selected corrected test. For example:

> Mauchly's test indicated that the assumption of sphericity was violated. Therefore, Greenhouse-Geisser-corrected results were interpreted. The effect of time remained statistically significant, *F*(1.84, 53.24) = 8.72, *p* = .001.

The numerical values in this example are illustrative. Actual values should always be taken from the statistical output.

## Example 3: Mixed-Design ANOVA with Three Repeated Measures

### Research Scenario

In a study of older women, researchers at Arizona State University were interested in whether high-intensity interval training (HIIT) would lead to greater improvements in cardiorespiratory fitness than a traditional steady-state elliptical training program.

To test their hypothesis, the researchers recruited women between the ages of 50 and 65 and randomly assigned participants to one of two conditions:

1. HIIT training exercise program
2. Steady-state training program

To assess cardiorespiratory fitness, the researchers measured VO₂ Max at three time points:

1. Baseline
2. Three months
3. Six months

### Design

The within-subjects factor is `time`, with three levels:

- Baseline
- Month 3
- Month 6

The between-subjects factor is `group`, with two levels:

- HIIT
- Steady-State

The dependent variable is VO₂ Max, measured in mL/kg/min.

### Hypothesized Means

| Group | Baseline | Month 3 | Month 6 |
|:---|---:|---:|---:|
| Steady-State | 35 | 37 | 39 |
| HIIT | 35 | 41 | 46 |

Both groups are expected to improve over time, but the HIIT group is expected to improve more rapidly.


```{r, echo=F}
set.seed(627)

n_group <- 30

steady_effect <- rnorm(n_group, 0, 3)
hiit_effect <- rnorm(n_group, 0, 3)

steady <- data.frame(
  id = 1:n_group,
  group = "Steady-State",
  baseline = 35 + steady_effect + rnorm(n_group, 0, 2.5),
  month3 = 37 + steady_effect + rnorm(n_group, 0, 2.5),
  month6 = 39 + steady_effect + rnorm(n_group, 0, 2.5)
)

hiit <- data.frame(
  id = (n_group + 1):(2 * n_group),
  group = "HIIT",
  baseline = 35 + hiit_effect + rnorm(n_group, 0, 2.5),
  month3 = 41 + hiit_effect + rnorm(n_group, 0, 2.5),
  month6 = 46 + hiit_effect + rnorm(n_group, 0, 2.5)
)

hiit_df <- bind_rows(steady, hiit) |>
  mutate(
    group = factor(group, levels = c("Steady-State", "HIIT"))
  )

head(hiit_df)
```

### Convert to Long Format

```{r, echo=F}
long_hiit_df <- hiit_df |>
  pivot_longer(
    cols = baseline:month6,
    names_to = "time",
    values_to = "vo2max"
  ) |>
  mutate(
    time = factor(
      time,
      levels = c("baseline", "month3", "month6"),
      labels = c("Baseline", "Month 3", "Month 6")
    )
  )

head(long_hiit_df)
```

### Examine Cell Means

```{r}
long_hiit_df |>
  group_by(group, time) |>
  summarise(
    n = n(),
    mean_vo2max = mean(vo2max),
    sd_vo2max = sd(vo2max),
    .groups = "drop"
  )
```

### Conduct the Mixed-Design ANOVA using afex 

```{r}
model_vo2 <- aov_ez(
  id = "id",
  dv = "vo2max",
  within = "time",
  between = "group",
  data = long_hiit_df
)

anova(model_vo2)
```
Again, there is a statistically signficant interaction indicating that the time effect varies by group. We can start with a visualization to get a sense of what that interaction might look like.

## Visualize the Interaction

```{r}
long_hiit_df |>
  group_by(group, time) |>
  summarise(
    mean_vo2 = mean(vo2max),
    .groups = "drop"
  ) |>
  ggplot(
    aes(
      x = time,
      y = mean_vo2,
      color = group,
      group = group
    )
  ) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    x = "Time",
    y = "Mean VO₂ Max",
    color = "Training Program"
  ) +
  theme_minimal()
```
## Interpreting the ANOVA results

The statistically significant F tests suggested that:

1. Main Effect of Time: VO₂ Max scores were different across time from baseline to Month 3 to Month 6, averaging across both training programs.
2. Main effect of Group: Participants in the HIIT and Steady-State programs had different $VO_2$ max scores averaging across all three measurement occasions.
3. Time-by-Group Interaction:  The pattern of change in VO₂ Max over time differed between the two exercise programs

The interaction is the most interesting and important effect, because the main hypothesis suggests that the effect of time on increased $VO_2$ Max is ***different*** depending on which group someone is in.

Put another way, everyone should show an increase, but that increase is moderated by, or depends on which exercise regimen they were assigned to.
    

### Evaluating Sphericity

Because this study contains three repeated measurements, sphericity must be considered. The `afex` output can provide corrected tests for within-subjects effects.

```{r}
summary(model_vo2)
```
These results suggested that there were no violations of the sphericity assumption, because the Mauchly Test was n.s.

If sphericity were violated, then the alternative results presented above could be used to draw study conclusions via the Greenhouse-Geisser or Huynh-Feldt corrected results. 

### Follow-Up Comparisons

If the interaction is statistically significant, follow-up analyses can identify where the groups differ.

#### Compare Groups at Each Time Point

```{r}
emmeans(
  model_vo2,
  pairwise ~ group | time)
```
At 3 and 6 months we see a group difference, but not at baseline.

#### Compare Time Points Within Each Group

```{r}
emmeans(
  model_vo2,
  pairwise ~ time | group)
```
These results suggest that we see differences across time points in both groups that are statistically significant. Both groups improved in their cardiorespiratory capacity. We can see that those differences across time points appear larger and that is what the interaction test tells us as well. 

# Analysis of Covariance

## What Is ANCOVA?

ANCOVA, or analysis of covariance, combines ANOVA and regression. It allows researchers to compare group means while statistically controlling for one or more continuous variables called covariates.

This is a type of ANOVA focused on mean group differences across levels of a factor, but we also introduce other variables (often continuous) to control for their possible effects on the outcome. 

The general ANCOVA model is:

$$
Y = \text{Group Effects} + \text{Covariate Effects} + \varepsilon
$$

or:

$$
Y = X\beta + C\beta_C + \varepsilon
$$

where:

- $X\beta$ represents the ANOVA portion of the model.
- $C\beta_C$ represents the regression portion of the model.
- $\varepsilon$ represents random error.

### Why Use a Covariate?

A covariate can:

- Explain variance in the dependent variable.
- Reduce unexplained error variance.
- Increase statistical precision.
- Adjust for preexisting differences among participants.

ANCOVA evaluates whether group differences remain after statistically controlling for the covariate.

### Interpreting ANCOVA

Unlike an unadjusted ANOVA, ANCOVA compares **adjusted means**, also called estimated marginal means.

Conceptually, ANCOVA asks:

> What would the group differences look like if all groups had the same value on the covariate?

As a result, effects that appear significant in an unadjusted ANOVA may become nonsignificant after an important covariate is included.

## Key Assumptions

Important ANCOVA assumptions include:

1. The covariate has an approximately linear relationship with the dependent variable.
2. The covariate is measured reliably.
3. The relationship between the covariate and outcome is similar across groups.
4. Observations are independent.
5. Residuals are approximately normally distributed.
6. Residual variances are reasonably similar across groups.
7. There are no highly influential outliers.
8. The covariate is measured before the outcome and is appropriate for adjustment.

The assumption that the regression relationship is the same across groups is called **homogeneity of regression slopes**. It can be evaluated by including factor-by-covariate interactions.

For example:

$$
Y =
A + B + C + A \times B + A \times C + B \times C
$$

where $C$ is the covariate.

A significant factor-by-covariate interaction suggests that the relationship between the covariate and the outcome differs across groups. In that situation, a single adjusted group comparison may not adequately summarize the data.

## Example 1: Mental Health Intervention for Latina Women

Suppose researchers evaluate a culturally tailored mental health intervention designed for Latina women experiencing psychological distress.

### Independent Variable

- Treatment condition:
  - Intervention
  - Control

### Dependent Variables

- Anxiety symptoms
- PTSD symptoms
- Depression symptoms

### Covariates

- Baseline anxiety
- Baseline PTSD
- Baseline depression

Participants complete measures before the intervention begins and again after the intervention concludes.

The primary research question is:

> Do women receiving the intervention report lower anxiety, PTSD, and depression at posttest than women in the control group, after controlling for baseline symptom levels?

The ANCOVA model for anxiety is:

$$
\text{Posttest Anxiety} =
\beta_0 +
\beta_1(\text{Treatment}) +
\beta_2(\text{Baseline Anxiety}) +
\varepsilon
$$

## ANCOVA and Multiple Regression

ANCOVA is a linear regression model that includes categorical predictors and continuous covariates. The R syntax for ANCOVA and multiple regression is therefore the same.

Both analyses can address whether group mean differences remain after adjusting for a covariate. In experimental designs, the covariate is often a baseline score on the outcome.

In this example, baseline anxiety is not necessarily a nuisance variable. It may improve precision by accounting for individual differences that existed before treatment began.

By controlling for baseline scores, researchers obtain a more precise estimate of the treatment effect. However, the interpretation depends on the study design and on whether the covariate was measured before treatment.

## Example 2: Generational Status and Breastfeeding Attitudes

Suppose researchers compare Latina mothers from three generational groups on attitudes toward breastfeeding. Some research suggests that acculturation among later-generation women may be associated with less positive views of breastfeeding and more favorable views of infant formula.

Family support of breastfeeding may also influence breastfeeding attitudes. Researchers therefore wish to determine whether generational differences remain after controlling for levels of family support.

### Independent Variable

Generational status:

- First-generation immigrant
- Second-generation
- Third-generation or later

### Dependent Variable

- Attitudes toward breastfeeding

### Covariate

- Family support

The research question is:

> Do breastfeeding attitudes differ across generational groups after controlling for family support for breastfeeding?

The ANCOVA model is:

$$
\text{Breastfeeding Attitudes} =
\beta_0 +
\beta_1(\text{Generation}) +
\beta_2(\text{Support}) +
\varepsilon
$$

In this example, support functions as an extraneous or explanatory variable because it may account for some differences in breastfeeding attitudes that are not attributable to generational status.

## Simulation Study for ANCOVA

### Hypothesis

First-generation women are expected to report more positive breastfeeding attitudes than second-generation and third-generation-or-later women.

Researchers also believe that family support for breastfeeding may explain some of the generational differences. They therefore conduct an ANCOVA to compare adjusted breastfeeding-attitude means while controlling for family support.

### Variables

#### Independent Variable

Generational status:

- First generation
- Second generation
- Third generation or later

#### Dependent Variable

The Iowa Infant Feeding Attitude Scale (IIFAS):

- A continuous numeric score.
- Higher scores indicate more positive attitudes toward breastfeeding.
- The scale is translated and back-translated for use with Spanish speakers.

#### Covariate

- Family support for breastfeeding

### Simulate Data

```{r}
library(tidyverse)

set.seed(627)

n_group <- 80

generation <- factor(
  rep(
    c(
      "First Generation",
      "Second Generation",
      "Third Generation+"
    ),
    each = n_group
  ),
  levels = c(
    "First Generation",
    "Second Generation",
    "Third Generation+"
  )
)

# Family support on a 1-to-10 scale.
family_support <- c(
  rnorm(n_group, mean = 8, sd = 1.2),
  rnorm(n_group, mean = 7, sd = 1.2),
  rnorm(n_group, mean = 6, sd = 1.2)
)

family_support <- pmin(pmax(family_support, 1), 10)

# IIFAS scores.
iifas <- 50 +
  c(
    rep(8, n_group),
    rep(4, n_group),
    rep(0, n_group)
  ) +
  2.5 * family_support +
  rnorm(3 * n_group, 0, 5)

iifas <- round(iifas)

bf_data <- data.frame(
  generation,
  family_support,
  iifas
)

head(bf_data)
```

### Examine Descriptive Statistics

```{r}
bf_data |>
  group_by(generation) |>
  summarise(
    n = n(),
    mean_iifas = mean(iifas),
    sd_iifas = sd(iifas),
    mean_support = mean(family_support),
    .groups = "drop"
  )
```

### Visualize the Data

```{r}
ggplot(
  bf_data,
  aes(
    x = generation,
    y = iifas,
    fill = generation
  )
) +
  geom_boxplot() +
  labs(
    x = "Generational Status",
    y = "IIFAS Score"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Unadjusted One-Way ANOVA

The unadjusted model compares mean IIFAS scores across generational groups without controlling for family support.

```{r}
bf_model <- aov(iifas ~ generation,
  data = bf_data)

summary(bf_model)

TukeyHSD(bf_model)
```
These results suggest that there is a generational difference in breasfeeding attitudes overall, then the folow up tests indicated that first generation mothers have more positive breastfeeding attitudes compared to both second and third-generation mothers. In addition, second-generation mothers have more positive breastfeeding attitudes when compared to third-generation mothers.

## ANCOVA Model: Do these differences remain after controlling for family support?

```{r}
bf_ancova <- lm(iifas ~ generation + family_support,
  data = bf_data)

summary(bf_ancova)
```
These results suggest that generational difference persist, even after controlling for family support

Can use these results (Based on regression) to examine ANOVA-like output using car package

```{r}

car::Anova(bf_ancova)

```

### Unadjusted and Adjusted Means

The following model produces unadjusted estimated means from the ANCOVA we just ran:

```{r}
library(emmeans)

emmeans(bf_model,~ generation)
```

The following model produces adjusted estimated marginal means, averaging over family support:

```{r}
adjusted_means <- emmeans(bf_ancova,~ generation)

adjusted_means
```

Adjusted pairwise comparisons can be obtained as follows:

```{r}
pairs(adjusted_means)
```
Again, we see that there are statistically signficant generational differences, even after controlling for family support.

## Testing Homogeneity of Regression Slopes

To evaluate whether the association between family support and IIFAS scores differs across generational groups, include the generation-by-family-support interaction:

```{r}
assumption_model <- lm(iifas ~ generation * family_support,
  data = bf_data)

car::Anova(assumption_model,type = 2)
```

A statistically significant `generation:family_support` interaction suggests that the slopes differ across generational groups. If this interaction is important, the assumption of homogeneous regression slopes is not supported, and the ANCOVA should not be summarized using one common adjusted group comparison. In this case the interaction was *not* statistically significant and so the assumption is met. 

