---
title: "Chapter 3: Data Cleaning and Getting into R"
format:
  html:
    page-layout: full
---

# Chapter Outline

1.  Intro to R
2.  Reading data into R
3.  Cleaning and checking data with respect to
    1.  Outliers
    2.  Missing data
    3.  Any miscoded or illogical scores/entries
4.  Recoding variables
    1.  Dummy coding nominal predictor variables
    2.  Reverse scoring any negative worded items
    3.  Transforming any variables that may have skew
    4.  Creating summary scores
    5.  Creating standardized variables
5.  Ensuring missing data flags are correct

## Introduction to R

R is a programming environment in which researchers can do a variety of different things. Most typically we will use R to:

1.Manage data sets 2.Clean, code data, create summary scores 3.Do statistical analyses like t-tests, regression, multilevel modeling, SEM 4.Create graphs, tables, and figures

But R can do a number of other tasks for us like reading OSF pages, scraping internet sties, renaming files, and converting files to PDFs.

Why R?

- It is Free!
- It is open-source which means that anyone can access, add to R and its numerous packages
- It uses a simple programming language developed by experts, you can see all source code (unlike some proprietary stats packages)
- This all means that there are new additions to R all the time to keep up with state-of-the-art changes in statistics

### Who will enjoy R the most?

- People who like programming!
- People who enjoy learning new code, exploring what R can do
- Those looking to streamline work time by using convenience features related to creating tables in APA style

Benefits of learning all that R can do for academic researchers

- Saves time and energy when creating reports
- Fewer mistakes transcribing output to tables and figures
- Can access anywhere (unlike networked software)
  - Makes it faster to analyze data
  - All analyses we want to do are possible with R

## What steps do you take to get started?

Download R from https://cran.r-project.org/

Download RStudio from https://posit.co/downloads

Find videos, tutorials, free books on GitHub, AIs are pretty good with R code

### What is the difference between R and RStudio?

R is the programming language in which we write code and execute commands. RStudio is a GUI (graphical user interface) that simplifies many of the basic tasks (e.g, reading in an SPSS file, installing packages, viewing plots) that we typically want to accomplish with R.

RStudio is what we will predominantly use in class and there are four sub-windows in R Studio:

- **Console** for running code immediately
- **Script editor** for saving code or running chunks of code
- **Environment pane** showing objects you’ve created
- **Files/Plots/Help** pane for outputs and documentation

Beyond the advantage of being able to more easily do basic tasks than with R, RStudio also has syntax highlighting, code suggestion/completion, spellcheck, and some smart coding routines

## Some Basic R Terminology

### Objects

***Objects*** store some kind of information, and have a name. These can be data sets, output files, or just simple numerical results.

- Frequently objects are created via the \<- or "assign" command, but can also use "="

```{r}
mean_y <- (9+8+4)/3 

# This "assigns" the mean of three numbers to "mean_y"

# If I 'print' the object, I get 7 or the mean of the numbers

mean_y

# Usually would use a function, however
# Let's pretend I have a *vector* of numbers

y <- c(9,8,4) # create object 'y' as a vector of three numbers
mean(y) # A built in function that just computes the mean of the y vector
        # Obviously also gives 7
        # The point here is just creating objects and using functions

# Often we are going to use pre-built functions, sometimes from baseR
# Often from packages

```

## Packages and Functions

There is Base R which has a number of **functions**

- Functions are like commands that have built in coded instructions to complete a task
- Some are pretty simple:
  - mean(var1) just computes the mean of var1
- Others are more complicated
  - lm for example lm(y \~ x1 +x2, data=newdata)
    - Is a mutiple regression "function" with many built in commands and routines

**Packages** are developed and distributed by programmers

- Often researchers who write functions (routines) to do some statistical analysis
- Or others: Creating PDFs, creating APA style tables, extracting results

Once you install a package using install.packages(), you then "load" the package using library()

- Note that while you only have to install packages once (unless you change computers), you have to load packages every time you restart R
- Also note that you can skip the 'install.packages()' command in RStudio (See above)
- As you will see -- in the scripts that I write -- I usually have load(haven) at the top along with other packages, that way when I open and run the scripts, that step happens first - If you try to use commands from a package that isn't loaded, you will get an error

## Reading Data Using Base R

Base R includes several functions for loading common file types.

Here we will read in a simple .csv file called "cleaning" which I just named because we're going to use it to demo how to check and clean some data. The data set is made up and just has Gender (coded Male or Female), four items assessing anxiety symptoms (anx1-anx4) and self-reported income in 1000s of dollar/year.

### CSV Files

```{r}
library(knitr)

df_csv <- read.csv("cleaning.csv") # Reads in the data file
df_csv                             # Print out the object, which is the dataframe

```

### SPSS Files

I often find that I am reading in .sav or SPSS files. In that case I use the haven package

Note two things here:

1.  If you don't have haven installed, need to do that (that is a "one time" thing),
2.  During each new R session, need to load the packages you plan to use.

```{r}
# install.packages("haven")   # Installs haven
library(haven)                # Loads haven
new_df <- read_sav("cleaning.sav")   # read_sav is the key haven command
                                     # "Cleaning.sav is the data set
new_df                               # Prints out the dataset
                                     # Usually I would use View for larger files
```

***NOTE*** that in the above 'tibble' there is additional information, for example the data type that was created. Gender is chr or character, while the other variables are dbl or double-numeric (having decimal points). May also see 'int' for integers

## Cleaning, Screening, and Dealing with Missing Data

Typically we organize and store data in some sort of spreadsheet. SPSS users for example might read in data from an excel spreadsheet or manually type in data.

Data in this instance refers to numeric values that represent things like participants:

- Age, gender, race-ethnicity
- Levels of stress, depression, motivation to exercise
- ID number identifiers for participants, families, schools, couples, neighborhoods

Regardless, data is often entered into a spreadsheet that really could use any software. Excel, notepad, SPSS, for example, and R can read in most any spreadsheet formats.

For new data, a wise initial step before getting into more sophisticated analyses is to use some basic descriptive statistics to screen one's data

## Initial Screening Checklist

The steps I often take when working with data for the first time:

1.  Read data in using base R or haven
2.  Use View(data) command to just look at the data like a spreadsheet a. Here I am looking to see that column names look correct, that values seem reasonable, just a quick peek
3.  Load the *psych* package and run describe(data) to get a comprehensive set of results that starts assisting with the identification of any data problems a. By looking a minimums and maximums, range, means
4.  Also can check for duplicate records
5.  Also check for outliers (see below)
6.  How much missing data are there, how will I avoid listwise deleting?

```{r}
#| eval: false
library(psych)
describe(df_csv)
```

```{r, echo=FALSE}
library(psych)
round(describe(df_csv), 2) |>
  knitr::kable()
```

## Checking the results for data problems

Some things to notice -

- Gender is a "character" variable that uses letters and not numbers

- The range for the anx items is 1 to 5 and so there is a data error for item anx3 -

  - 33 is an invalid number, it could have been double entered

- Income may have an outlier

  - This is a plausible value, just extreme

  - The presence of that score contributes to some mild skew, and nonnormality

  - This is also evident as the means and medians are quite different

## Making changes to the dataframe

- This will fix the 33 value and recode to a 3 (assuming it was double entered)
- Then we will "winsorize" income to see if we can reduce skew
  - Options for dealing with outliers
    - Delete the case (loses information)
    - Transformations of variables to deal with skew (affects all scores)
    - Winsorizing or bringing a value up/down while retaining ordering of scores
- Here I make those changes and look at the data again
  - Note that this tiny toy dataset is too small for Winsorizing in the typical sense to make a big difference, so I also did a more extreme rescaling
  - Note also that Winsorizing in the DescTools package won't work with missing data

```{r}
# Recoding the 33 to a 3
df_csv$anx3[df_csv$anx3 == 33] <- 3
# Now will winsorize income 
#install.packages("DescTools")
library(DescTools)
df_csv$income_w <- Winsorize(df_csv$income) # This won't do much, small sample
df_csv$income_w2 <- ifelse(df_csv$income > 100, 100, df_csv$income) # more dramatic

```

```{r, echo=FALSE}
round(describe(df_csv), 2) |>
  knitr::kable()
```

### Other Tranformations

- Let's say that anxiety item 4 was in need of reverse scoring.

- This often arises in cases where items have different "valence" in terms of positive or negative valence.

  - For example: anx1 asked participants "How often have you experienced overwhelming worry over the past week?" with response options of: Never, Rarely, Sometime, Often, Very Often
  - anx4 asked "Have you been able to easily relax in the past week?".
  - Note that a score of '5' for one item would indicate frequent experiences of anxiety, but a low anxiety on the other.
  - In cases like this, items that are counter to the desired valence - in this case greater frequency of anxiety - would be reverse coded so that an overall summary score doesn't combine high and low anxiety across the items

  ```{R}
  df_csv$anx4 = 5-df_csv$anx4 # This converts item 4 scores from: 4 --> 1, 3 --> 2, etc

  # After reverse coding, I want to create a summary score

  df_csv$anxiety_mean <- rowMeans(df_csv[, c("anx1", "anx2", "anx3", "anx4")],
    na.rm = TRUE)

  # This creates a new variable 'anxiety_mean' in the df_csv dataframe, it creates means for  rows (people) across the columns anx1, anx2, etc. na.rm = TRUE keeps participants
  # even if they don't have scores for all of the items (avoids listwise deleting)

  # The resulting dataset now has a summary score for anxiety_mean that is 
  # scores properly


  ```

  ```{r, echo=F}
  library(psych)
  round(describe(df_csv), 2) |>
    knitr::kable()
  ```

  ::: callout-warning
  ## What Are Outliers?
  :::

An **outlier** is an observation that is unusually distant from the bulk of the data — either extremely high, extremely low, or both (in multivariate settings). Outliers may reflect:

- **Data entry errors** (e.g., age coded as 999)
- **Measurement error** (e.g., equipment malfunction)
- **True but rare observations** (e.g., a legitimately high income in a low-income sample)
- **Violations of study protocol** (e.g., participant who didn't follow instructions)

### How to Detect Outliers

**Univariate (single variable):** -

- **Visual:** Boxplot, histogram, scatterplot — look for points far from the main cluster

- **IQR method:** Mark as outlier if value \< Q1 − 1.5×IQR or \> Q3 + 1.5×IQR -

- **Z-score method:** Mark as outlier if \|z\| \> 3 (more conservative: \|z\| \> 2.5) -

- **Range checks:** Compare to known valid ranges (e.g., age should be 0–120)

**Multivariate (multiple variables):**

- **Mahalanobis distance:** Measures how far a case is from the centroid of all variables combined

- **Cook's distance:** Identifies cases with high influence on regression coefficients

- **Leverage + residuals:** Cases with high leverage and large residuals are flagged

**R Code Example:**

```{r}
#| label: outlier-detection

library(dplyr)

# IQR method
df <- tibble(age = c(25, 30, 28, 32, 999, 27, 29))  # 999 is an outlier

Q1 <- quantile(df$age, 0.25)
Q3 <- quantile(df$age, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

outliers <- df %>%
  filter(age < lower_bound | age > upper_bound)

# Z-score method
df <- df %>%
  mutate(
    z_age = scale(age)[, 1],
    is_outlier = abs(z_age) > 3
  )


```

:::: \## Coding Categorical Variables

- **Nominal**: no inherent order (e.g., race/ethnicity, program site, referral source).
- **Ordered categorical**: ordered categories without equal spacing assumed (e.g., Likert items, severity stages: mild/moderate/severe).
- **Dummy coding**: reference group = 0, comparison groups = 1; coefficients interpreted relative to the reference.
- **Effect coding**: groups coded -1/1; coefficients interpreted relative to the grand mean.
- Note that we will get into coding specifics in R when we run analyses...

::: callout-warning
## Common Pitfall

Numeric-looking codes (`1 = Male, 2 = Female`) get accidentally treated as continuous in R if not explicitly converted to a factor — this silently corrupts downstream models. Also, there are many good reasons for why binary variables should be recoded to 0/1. For example, got divorced = 1, did not get divorced = 0 as opposed to 1 and 2.

- Easier interpretation of intercepts in regressions
- Easier interpretation when doing product terms for interactions
- Mean is actually meaningful (will be percentage of participants in 1 category)
:::

## R Code: Factors and Recoding

```{r}
#| label: categorical-coding

df <- tibble(
  client_id = 1:6,
  sex = c(1, 2, 2, 1, 1, 2),
  severity = c("mild", "severe", "moderate", "mild", "severe", "moderate")
)

# Convert to factor with meaningful labels
df <- df %>%
  mutate(
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    severity = factor(severity,
                       levels = c("mild", "moderate", "severe"),
                       ordered = TRUE)
  )

# Set/change the reference level for regression
df$sex <- relevel(df$sex, ref = "Male")

# Quick check
levels(df$sex)
table(df$sex, useNA = "always")
str(df)
```

::: callout-important
## Discussion Questions

We need to be really careful with nominal IVs. Say for example I want to see if there is greater participation in blood pressure screening at Urban Health Clinics, Rural Clinics, or via TeleHealth visits (using home BP monitor)

I would most likely have a variable titled "Clinic Type" that has values of 1, 2, and 3 that represent Urban, Rural, and Telehealth, respectively

In an ANOVA, that isn't much of a problem because the RQ is: Are there differences in participation across three clinics? In Multiple regression, however, we are asking: Is a one level *increase* in clinic type (meaningless) associated with greater BP screening participation

In a case like this we need to ***dummy code*** site type. Will come back to this later...
:::

## Missing Data

### Will review later but quickly...

- **MCAR** (Missing Completely at Random) — missingness unrelated to any variable
- **MAR** (Missing at Random) — missingness related to observed variables
- **MNAR** (Missing Not at Random) — missingness related to the unobserved value itself

::: callout-note
**Decide your missing-data policy before analysis**, and document it — don't let listwise deletion happen silently by default.

Note that in R, missing data gets referred to as NA
:::

## R Code: Screening for Data Quality Issues

```{r}
#| label: missingness-screening

library(naniar)  # A package with commands/functions to examine missing data

# Example dataset with missing data and outliers

df_raw <- tibble(
  client_id = c(1, 2, 2, 3, 4, 5),
  age = c(34, 150, 150, -2, 45, NA),
  program_site = c("Urban", "Rural", "Rural", "Urban", NA, "Telehealth")
)

# Missingness summary by variable
miss_var_summary(df_raw)

```

## Data Structures: Wide vs. Long

**Wide format** — one row per subject, repeated measures spread across columns:

| id  | score_t1 | score_t2 | score_t3 |
|-----|----------|----------|----------|
| 1   | 12       | 15       | 18       |
| 2   | 9        | 11       | 10       |

**Long format** — one row per subject-time observation:

| id  | time | score |
|-----|------|-------|
| 1   | t1   | 12    |
| 1   | t2   | 15    |
| 1   | t3   | 18    |
| 2   | t1   | 9     |
| 2   | t2   | 11    |
| 2   | t3   | 10    |

::: callout-note
`ggplot2`,

Most cross-sectional data or SEM-type data are going to be "wide" format - so there is one row per participant, and each column represents a different variables. If the data are repeated measures/longitudinal, each variable would just be a score at a particular time point

Many "mixed-effects" models (for example the lme4 package) will read in long-format data frames. So there will be multiple rows per participant and there often need to be more than one ID variable. Then there are columns for each variable and the rows represent different scores that are "nested" in teach participant.
:::

## R Code: Reshaping Data

R uses the work "reshaping", other terms include "turning" the data or "reformatting"

```{r}
#| label: Converting wide data to long

library(tidyr)
library(dplyr)

# First creating a "toy" data set to play with 
# Note a tibble is just a term for a simple rows/columns entry
# This is just creating a quick data fram with ID and two rows
# Then score_t1 = the variable t1_anx is anxiety @ time 1
# Then t_2 and t_3 are for time 2 and 3 - so longitudinal
# Then the parentheses hold made up scores
df_wide <- tibble(
  id = 1:2,
  t1_anx = c(12, 9),
  t2_anx = c(15, 11),
  t3_anx = c(18, 10))

# Wide -> Long
df_long <- df_wide %>%
  pivot_longer(
    cols = starts_with("t"),
    names_to = "time",
    names_prefix = "anxiety",
    values_to = "score")

df_long

# Long -> Wide 

df_wide2 <- df_long %>%
  pivot_wider(
    names_from = time,
    names_prefix = "anxiety",
    values_from = score
  )

df_wide2


```

You can see from these that the scores that were horizontal "turn" or are reshaped to vertical and vice versa :::

## More on Getting Data into R

Most of my experience reading in data has to do with bringing data in from some sort of spreadsheet like in an excel or SPSS file.

There are **BaseR** commands to read in .txt files like:

- df \<- read.csv("myfile.csv") : For comma separated files (csv)
- df \<- read.delim("myfile.txt") : For .txt or text files with tab delimiters (spacing)
- df \<- read.table("myfile.txt", header = TRUE) : For generic text files
- header = TRUE means that the first row includes column (variable) names

# Packages with functions to read in data, often from specific software

- Tidyverse has a package called "readr" - name of the dataset goes in the ()

  - read_csv() CSV
  - read_tsv() Tab-delimited
  - read_delim() User-specified delimiter
  - read_table() Fixed-width text tables
  - read_fwf() Fixed-width files
  - read_rds() will explain that later...

Notice that there are very **similar** commands/functions here, the . and \_ differentiate the package and what these do... easy to get confused.

- The haven Package - note, no caps in 'haven'

  - read_sav() \# SPSS
  - read_por() \# SPSS portable - usually has been converted first
  - read_dta() \# Stata
  - read_sas() \# SAS
  - read_xpt() \# SAS transport

- You may also come across the foreign package, older probably less used

  - read.spss() \# SPSS
  - read.dta() \# Stata

## Data Frames

Datasets are typically stored in ***data frames***. R will refer to data frames has having Rows and Columns.

Typically in a wide-format dataset, each person/case/subject/participant/unit will represent a row, while each variable will be in a column. It is helpful to remember that terminology because we will sometimes reference columns and rows.

## Saving out dataframes for later use in R

Typically I might read in a .sav file that I or a colleague created.

```{r}
library(haven)

my_data <- read_sav("C:\\Users\\ajsup\\Desktop\\Hmong Data File.sav")

# Then lets say I do some recoding

my_data$gender <- factor(my_data$Gender,
                    levels = c(1, 2),
                    labels = c("Female", "Male")) # All this does is create a factor
                                                  # Variable with character values, not numbers

# Then I save out an RDS file which is an R dataframe

saveRDS(my_data, "C:/Users/ajsup/Desktop/DCYA_clean.rds")

# Next time I go into R, I don't have to reload haven, read in the file, etc

my_data <- readRDS("C:/Users/ajsup/Desktop/DCYA_clean.rds")

prop.table(table(my_data$Gender))
prop.table(table(my_data$gender))

```
