---
title: "Chapter 1: Data Foundations for HHS Research"
author: "Dr. Andy Supple"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
editor: source
execute:
  eval: false
  warning: false
  message: false
---



------------------------------------------------------------------------

# Data Structures: Wide vs. Long

**Wide format** — one row per subject, repeated measures spread across columns:

| id  | score_t1 | score_t2 | score_t3 |
|-----|----------|----------|----------|
| 1   | 12       | 15       | 18       |
| 2   | 9        | 11       | 10       |

**Long format** — one row per subject-time observation:

| id  | time | score |
|-----|------|-------|
| 1   | t1   | 12    |
| 1   | t2   | 15    |
| 1   | t3   | 18    |
| 2   | t1   | 9     |
| 2   | t2   | 11    |
| 2   | t3   | 10    |

::: callout-note
`ggplot2`, mixed-effects models (`lme4`), and growth curve models expect long format. Wide format is often how data arrive from REDCap/Qualtrics exports, so reshaping is a routine first step, not an edge case.
:::

## R Code: Reshaping Data

```{r}
#| label: setup
#| eval: false

library(tidyverse)
library(here)
library(haven)
library(naniar)
library(janitor)
```

```{r}
#| label: wide-long-example

library(tidyr)
library(dplyr)

# Example wide data
df_wide <- tibble(
  id = 1:2,
  score_t1 = c(12, 9),
  score_t2 = c(15, 11),
  score_t3 = c(18, 10)
)

# Wide -> Long
df_long <- df_wide %>%
  pivot_longer(
    cols = starts_with("score_t"),
    names_to = "time",
    names_prefix = "score_t",
    values_to = "score"
  )

df_long

# Long -> Wide (reverse, e.g., for a codebook table)
df_wide2 <- df_long %>%
  pivot_wider(
    names_from = time,
    names_prefix = "score_t",
    values_from = score
  )

df_wide2
```


------------------------------------------------------------------------

# Coding Categorical Variables

- **Nominal**: no inherent order (e.g., race/ethnicity, program site, referral source).
- **Ordered categorical**: ordered categories without equal spacing assumed (e.g., Likert items, severity stages: mild/moderate/severe).
- **Dummy coding**: reference group = 0, comparison groups = 1; coefficients interpreted relative to the reference.
- **Effect coding**: groups coded -1/1; coefficients interpreted relative to the grand mean.
-  Note that we will get into coding specifics in R when we run analyses...

::: callout-warning
## Common Pitfall

Numeric-looking codes (`1 = Male, 2 = Female`) get accidentally treated as continuous in R if not explicitly converted to a factor — this silently corrupts downstream models.
:::

## R Code: Factors and Recoding

```{r}
#| label: categorical-coding

df <- tibble(
  client_id = 1:6,
  sex = c(1, 2, 2, 1, 1, 2),
  severity = c("mild", "severe", "moderate", "mild", "severe", "moderate")
)

# Convert to factor with meaningful labels
df <- df %>%
  mutate(
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    severity = factor(severity,
                       levels = c("mild", "moderate", "severe"),
                       ordered = TRUE)
  )

# Set/change the reference level for regression
df$sex <- relevel(df$sex, ref = "Male")

# Quick check
levels(df$sex)
table(df$sex, useNA = "always")
str(df)
```

::: callout-important
## Discussion Questions

We need to be really careful with nominal IVs. Say for example I want to see if there is greater participation in blood pressure screening at Urban Health Clinics, Rural Clinics, or via TeleHealth visits (using home BP monitor)

I would most likely have a variable titled "Clinic Type" that has values of 1, 2, and 3 that represent Urban, Rural, and Telehealth, respectively

In an ANOVA, that isn't much of a problem because the RQ is: Are there differences in participation across three clinics? In Multiple regression, however, we are asking: Is a one level *increase* in clinic type (meaningless) associated with greater BP screening participation

In a case like this we need to ***dummy code*** site type. Will come back to this later...


:::

------------------------------------------------------------------------

# Cleaning, Screening, and Dealing with Missing Data

Typically we organize and store data in some sort of spreadsheet. SPSS uses for example might read in data from an excel spreadshoot or manually type in data. 

Data in this instance refers to numeric values that represent things like participants:
-Age, gender, race-ethnicity
-Levels of things like stress, depression, motivation to excercise
-ID number identifiers for participants, families, schools, couples, neighborhoods

Regardless, data is often entered into a spreadsheet that really could use any software.

R will read in most any spreadsheet formats

For new data, a wise initial step before getting into more sophisticated anlayses is to use some basic descriptive statistics to screen one's data

## Initial Screening Checklist

-Use software to run some basic descriptive statistics
-My preferred workflow is to open data in R, then load the *psych* package to run descriptives. That produces a comprehensive set of results that start assisting

-One should check the range of values for each variable to see if any out of range values (age = 222) appear
-Any illogical inconsistences, duplicate records, or outliers


Range checks (e.g., age = 150, negative service counts)
- Impossible/logical inconsistencies (e.g., discharge date before admission date)
- Duplicate records (same client ID appearing twice)
- Outlier flags (visual and/or IQR-based)

::: callout-warning

### What Are Outliers?

An **outlier** is an observation that is unusually distant from the bulk of the data — either extremely high, extremely low, or both (in multivariate settings). Outliers may reflect:

- **Data entry errors** (e.g., age coded as 999)
- **Measurement error** (e.g., equipment malfunction)
- **True but rare observations** (e.g., a legitimately high income in a low-income sample)
- **Violations of study protocol** (e.g., participant who didn't follow instructions)

### How to Detect Outliers

**Univariate (single variable):**
- **Visual:** Boxplot, histogram, scatterplot — look for points far from the main cluster
- **IQR method:** Mark as outlier if value < Q1 − 1.5×IQR or > Q3 + 1.5×IQR
- **Z-score method:** Mark as outlier if |z| > 3 (more conservative: |z| > 2.5)
- **Range checks:** Compare to known valid ranges (e.g., age should be 0–120)

**Multivariate (multiple variables):**
- **Mahalanobis distance:** Measures how far a case is from the centroid of all variables combined
- **Cook's distance:** Identifies cases with high influence on regression coefficients
- **Leverage + residuals:** Cases with high leverage and large residuals are flagged

**R Code Example:**

```{r}
#| label: outlier-detection

library(dplyr)

# IQR method
df <- tibble(age = c(25, 30, 28, 32, 999, 27, 29))  # 999 is an outlier

Q1 <- quantile(df$age, 0.25)
Q3 <- quantile(df$age, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

outliers <- df %>%
  filter(age < lower_bound | age > upper_bound)

# Z-score method
df <- df %>%
  mutate(
    z_age = scale(age)[, 1],
    is_outlier = abs(z_age) > 3
  )

# Mahalanobis distance (multivariate)
library(robust)
mahal <- mahalanobis(df[, c("age", "income")], 
                      center = colMeans(df[, c("age", "income")]),
                      cov = cov(df[, c("age", "income")]))
outlier_cases <- which(mahal > qchisq(0.99, df = 2))  # 99th percentile threshold

## Missing Data Mechanisms (Conceptual)

- **MCAR** (Missing Completely at Random) — missingness unrelated to any variable
- **MAR** (Missing at Random) — missingness related to observed variables
- **MNAR** (Missing Not at Random) — missingness related to the unobserved value itself

::: callout-note
**Decide your missing-data policy before analysis**, and document it — don't let listwise deletion happen silently by default.
:::

## R Code: Screening for Data Quality Issues

```{r}
#| label: missingness-screening

library(naniar)
library(janitor)

# Example dataset with issues
df_raw <- tibble(
  client_id = c(1, 2, 2, 3, 4, 5),
  age = c(34, 150, 150, -2, 45, NA),
  program_site = c("Urban", "Rural", "Rural", "Urban", NA, "Telehealth")
)

# Visualize missingness
vis_miss(df_raw)

# Missingness summary by variable
miss_var_summary(df_raw)

# Quick categorical screening
tabyl(df_raw, program_site)

# Duplicate check
df_raw %>% get_dupes(client_id)

# Range check example
df_raw %>% filter(age < 0 | age > 110)
```

::: callout-important
## Discussion Questions

10. Describe a plausible MAR (Missing at Random) scenario in an HHS client survey. How would you distinguish it from MNAR?
11. Run `vis_miss()` on a provided dataset. Write two sentences summarizing the missingness pattern you observe.
:::

------------------------------------------------------------------------

# Creating an Analysis-Ready Dataset

- **Analytic sample ≠ raw dataset.** Define inclusion/exclusion criteria explicitly in code (e.g., "exclude records missing baseline score," "restrict to clients enrolled ≥ 30 days").
- **Codebook as a deliverable**: variable name, label, type, valid values, missing codes — maintained alongside the dataset, not reconstructed from memory later.
- **Merging/joining** multiple sources: identify join keys, decide one-to-one vs. one-to-many, and check row counts before/after joins to catch duplication errors.
- **File pipeline discipline**: `raw/` → `cleaned/` → `analytic/` — never overwrite the raw file.

## R Code: Building the Analytic Sample

```{r}
#| label: analytic-sample

# Example: cleaned dataset already exists as df_cleaned
df_cleaned <- tibble(
  client_id = 1:6,
  baseline_score = c(10, NA, 8, 12, 9, 11),
  enrollment_days = c(45, 20, 60, 15, 90, 30)
)

# Define analytic sample transparently
df_analytic <- df_cleaned %>%
  filter(!is.na(baseline_score), enrollment_days >= 30)

df_analytic

# Join example with row-count sanity check
df_survey <- tibble(client_id = 1:5, satisfaction = c(4,5,3,4,5))
df_admin  <- tibble(client_id = 1:5, cost = c(100,200,150,120,300))

nrow(df_survey)
df_merged <- df_survey %>%
  left_join(df_admin, by = "client_id")
nrow(df_merged)  # should match df_survey unless intentional fan-out

# Save each stage distinctly (paths illustrative)
# saveRDS(df_cleaned, here("data", "cleaned", "clean_v1.rds"))
# saveRDS(df_analytic, here("data", "analytic", "analytic_v1.rds"))
```

::: callout-important
## Discussion Questions

12. Write, in plain language, the inclusion/exclusion criteria you would use to define an analytic sample for a study of first-year program dropout. Then translate that into a `dplyr::filter()` statement.
13. Why is it risky to overwrite a raw data file with cleaned data, even if you're confident in your cleaning code?
:::

------------------------------------------------------------------------

# R Workflows: Import, Clean, Label, Recode (Worked Example)

```{r}
#| label: full-workflow-example
#| eval: false

library(readr)
library(haven)
library(dplyr)
library(forcats)

# 1. IMPORT
raw <- read_csv("data/raw/survey_extract.csv")
# or, for SPSS/Stata files common in HHS agency data:
# raw <- haven::read_sav("data/raw/agency_data.sav")

# 2. INITIAL SCREEN
glimpse(raw)
summary(raw)

# 3. CLEAN
clean <- raw %>%
  distinct(client_id, .keep_all = TRUE) %>%          # remove duplicates
  filter(age >= 0, age <= 110) %>%                    # range check
  mutate(across(where(is.character), str_trim))       # trim whitespace

# 4. RECODE & LABEL
clean <- clean %>%
  mutate(
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    race_eth = fct_recode(race_eth,
      "White" = "1", "Black/African American" = "2",
      "Hispanic/Latino" = "3", "Other" = "4"
    ),
    severity = case_when(
      phq9_total <= 4  ~ "Minimal",
      phq9_total <= 9  ~ "Mild",
      phq9_total <= 14 ~ "Moderate",
      phq9_total <= 19 ~ "Moderately Severe",
      phq9_total >= 20 ~ "Severe",
      TRUE ~ NA_character_
    ),
    severity = factor(severity, levels = c("Minimal","Mild","Moderate",
                                            "Moderately Severe","Severe"),
                       ordered = TRUE)
  )

# 5. EXPORT ANALYSIS-READY FILE
write_csv(clean, "data/analytic/phq9_analytic.csv")
saveRDS(clean, "data/analytic/phq9_analytic.rds")
```

::: callout-important
## Exercise

14. Using the worked example script as a template, import a small provided dataset and produce a labeled, recoded analytic file. Submit both the script and the output codebook.
:::

------------------------------------------------------------------------

# Why Reproducibility Matters

- Reproducibility underpins **credibility** in HHS policy research — results that inform funding or program decisions must be independently verifiable.
- Practical habits: version control (Git basics), self-documenting scripts (comments explaining *why*, not just *what*), relative file paths (`here::here()`), recording `sessionInfo()` for package versions.
- Every cleaning decision (exclusions, recodes, missing-data handling) should live in a script — never in manual spreadsheet edits.
- **Bridge to later chapters**: the analytic dataset built here becomes the shared foundation for all subsequent hypothesis-testing chapters. If this foundation is undocumented or irreproducible, every downstream p-value is suspect.

## R Code: Reproducibility Habits

```{r}
#| label: reproducibility
#| eval: false

# Document environment for reproducibility
sessionInfo()

# Use project-relative paths
library(here)
df <- read_csv(here("data", "raw", "survey_extract.csv"))
```

::: callout-important
## Discussion Question

15. Reflect: has a research finding you've encountered (in class, in practice, or in the news) been difficult to trust because the underlying data process wasn't transparent? What would reproducibility have changed?
:::

------------------------------------------------------------------------

# Chapter Assessment

::: callout-tip
## Applied Assignment

Provide students a messy raw CSV (duplicates, out-of-range values, unlabeled categorical codes, some missingness) and require them to submit:

1.  A cleaning script
2.  A codebook
3.  A brief write-up justifying their inclusion criteria and missing-data decisions

This establishes the habits this chapter is built to instill before any hypothesis testing begins.
:::
