---
title: "Multiple Regression"
---

## Multiple Regression

Note that sometimes there is "regression" or bivariate regression which is not really different from a correlation, so unlikely you will ever run one of those

It is a helpful starting point to understand a basic formula and a scatterplot for how the regression line is estimated

### Modeling Relationships Between Variables

Goals are to link variables together: Correlations, regression, multiple regression In most of these we are discussing continous variables - particulaly for the outcomes -We will discuss **logistic** regression later which models binary outcomes with extensions to **multinomial** logistic regression for nominal outcomes

> **Understand • Model • Predict**

::: {.callout-note title="What Is It?" icon="false"}
Regression Analysis is a statistical technique used to model and analyze the relationship between a **dependent variable (Y)** and one or more **independent variables (X)**.

**Goal:** Predict or explain the value of Y based on the value(s) of X.
:::

Sometimes the goal *could* be to identify variance explained (R^2^), but for most HHS-type research purposes we are interested in questions like:

- What factors predict increased depressive symptoms?

- How does amount of exercise during a particular week predict positive mood **controlling for** life stressors, social support, and sleep?

- Do neighborhood effects influence health intentions above and beyond **social determinants** and demographic characterists (e.g., age, race-ethnicity, gender, socioeconomic status)\>

So our goals often might have to do with:

1.  Which is the most important predictor among many
2.  Increasing internal validity to rule out spurious effects
    1.  i.e, if we argue that case for causality/internal validity is improved if we establish
        1.  Association between X and Y
        2.  Temporal precedence where X preceds Y in time
        3.  Nonspurious associations - ruling out other possible causes
3.  Do social factors predict above and beyond background characteristics?
4.  More complicated questions related to mediation and moderation

------------------------------------------------------------------------

### How It Works

:::::: {layout-ncol="3"}
::: {.callout-tip title="Input"}
Independent Variable(s)

$$X_1, X_2, \dots, X_p$$
:::

::: {.callout-tip title="Regression Model"}
Estimates the relationship between the X(s) and Y
:::

::: {.callout-tip title="Output"}
Dependent Variable (Y)

Predicted $\hat{Y}$
:::
::::::

::: {.callout-important title="Key Idea" icon="false"}
Regression finds the **best-fitting line (or surface)** that minimizes the difference between observed values and predicted values.
:::

------------------------------------------------------------------------

### When to Use?

- When you want to predict a **continuous outcome**
- When examining the **relationship between variables**
- When **forecasting or trend analysis** is required
- When studying the **impact of one or more factors** on an outcome
- **Examples:** Sales forecasting, price prediction, exam score prediction, demand analysis, growth modeling

------------------------------------------------------------------------

### Types of Regression

| \# | Type | Description | Model |
|-----------------|-----------------|-----------------------|-----------------|
| 1 | **Simple Linear Regression** | One independent variable | $Y = \beta_0 + \beta_1 X + \varepsilon$ |
| 2 | **Multiple Linear Regression** | Two or more independent variables | $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_p X_p + \varepsilon$ |
| 3 | **Polynomial Regression** | Relationship modeled as an nth-degree polynomial | $Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \dots + \beta_n X^n + \varepsilon$ |
| 4 | **Logistic Regression** | Used when the dependent variable is categorical (0 or 1) | $\log\left(\dfrac{p}{1-p}\right) = \beta_0 + \beta_1 X + \dots + \beta_p X_p$ |

------------------------------------------------------------------------
## Worked Examples: Simple Linear Regression 

For a visual, let's run two bivariate (simple linear) regressions using simulated data on womens' self-reported relationship experiences and life satisfaction.

- **Example 1:** Marital satisfaction → Life satisfaction (expected **positive** association)
- **Example 2:** Partner conflict → Life satisfaction (expected **negative** association)



```{r setup, include=FALSE}
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
library(tidyverse)
library(broom)
set.seed(147)
```

### Simulating the Data

```{r simulate-data}
n <- 200

# Example 1: Marital satisfaction (X) -> Life satisfaction (Y), positive relationship
marital_satisfaction <- rnorm(n, mean = 5, sd = 1.2)
life_satisfaction_1  <- 2 + 0.65 * marital_satisfaction + rnorm(n, mean = 0, sd = 1)

# Example 2: Partner conflict (X) -> Life satisfaction (Y), negative relationship
partner_conflict     <- rnorm(n, mean = 4, sd = 1.5)
life_satisfaction_2  <- 8 - 0.55 * partner_conflict + rnorm(n, mean = 0, sd = 1)

df <- tibble(
  id                    = 1:n,
  marital_satisfaction  = marital_satisfaction,
  life_satisfaction_1   = life_satisfaction_1,
  partner_conflict      = partner_conflict,
  life_satisfaction_2   = life_satisfaction_2
)

head(df)
```

---

### Example 1: Marital Satisfaction Predicting Life Satisfaction

#### The Model

Following the general simple linear regression form from @sec-regression-overview:

$$
Y = \beta_0 + \beta_1 X + \varepsilon
$$

Mapped onto this example:

$$
\text{Life Satisfaction} = \beta_0 + \beta_1(\text{Marital Satisfaction}) + \varepsilon
$$

#### Running the Regression

```{r model-1}
model_1 <- lm(life_satisfaction_1 ~ marital_satisfaction, data = df)
summary(model_1)
```

```{r model-1-tidy}
tidy(model_1)
glance(model_1) |> select(r.squared, adj.r.squared, sigma)
```

#### Plotting the Regression Line

```{r plot-1, fig.cap="Marital satisfaction predicting life satisfaction (positive association)"}
ggplot(df, aes(x = marital_satisfaction, y = life_satisfaction_1)) +
  geom_point(color = "#2c3e91", alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "#c0392b", linewidth = 1) +
  labs(
    title = "Marital Satisfaction and Life Satisfaction",
    subtitle = expression(paste("Model: ", Y == beta[0] + beta[1] * X + epsilon)),
    x = "Marital Satisfaction (X)",
    y = "Life Satisfaction (Y)"
  ) +
  theme_minimal(base_size = 13)
```

#### Mapping to the Infographic Figure

| Infographic Element | This Example |
|---|---|
| **Input** ($X_1, X_2, \dots, X_p$) | Marital satisfaction scores |
| **Regression Model** (estimates relationship) | `lm(life_satisfaction_1 ~ marital_satisfaction)` |
| **Output** (Predicted $\hat{Y}$) | Predicted life satisfaction values (the red fitted line) |
| **Best-fitting line** | `geom_smooth(method = "lm")` — minimizes squared residuals |
| **$\beta_1$ (slope)** | `r round(coef(model_1)[2], 2)` — for each 1-unit increase in marital satisfaction, life satisfaction increases by this amount |
| **$\beta_0$ (intercept)** | `r round(coef(model_1)[1], 2)` — predicted life satisfaction when marital satisfaction = 0 |
| **R²** | `r round(summary(model_1)$r.squared, 3)` — proportion of variance in life satisfaction explained by marital satisfaction |

---

### Example 2: Partner Conflict Predicting Life Satisfaction

#### The Model

$$
\text{Life Satisfaction} = \beta_0 + \beta_1(\text{Partner Conflict}) + \varepsilon
$$

Here we expect $\beta_1 < 0$ — a **negative** association.

#### Running the Regression

```{r model-2}
model_2 <- lm(life_satisfaction_2 ~ partner_conflict, data = df)
summary(model_2)
```

```{r model-2-tidy}
tidy(model_2)
glance(model_2) |> select(r.squared, adj.r.squared, sigma)
```

#### Plotting the Regression Line

```{r plot-2, fig.cap="Partner conflict predicting life satisfaction (negative association)"}
ggplot(df, aes(x = partner_conflict, y = life_satisfaction_2)) +
  geom_point(color = "#8e44ad", alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "#c0392b", linewidth = 1) +
  labs(
    title = "Partner Conflict and Life Satisfaction",
    subtitle = expression(paste("Model: ", Y == beta[0] + beta[1] * X + epsilon)),
    x = "Partner Conflict (X)",
    y = "Life Satisfaction (Y)"
  ) +
  theme_minimal(base_size = 13)
```

#### Mapping to the Infographic Figure

| Infographic Element | This Example |
|---|---|
| **Input** ($X_1, X_2, \dots, X_p$) | Partner conflict scores |
| **Regression Model** | `lm(life_satisfaction_2 ~ partner_conflict)` |
| **Output** (Predicted $\hat{Y}$) | Predicted life satisfaction values (downward-sloping red line) |
| **$\beta_1$ (slope)** | `r round(coef(model_2)[2], 2)` — **negative** slope: each 1-unit increase in conflict is associated with a *decrease* in life satisfaction |
| **$\beta_0$ (intercept)** | `r round(coef(model_2)[1], 2)` |
| **R²** | `r round(summary(model_2)$r.squared, 3)` |

---

### Side-by-Side Comparison

```{r combined-plot, fig.cap="Comparing positive vs. negative bivariate associations with life satisfaction", fig.width=10, fig.height=5}
p1 <- ggplot(df, aes(x = marital_satisfaction, y = life_satisfaction_1)) +
  geom_point(color = "#2c3e91", alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "#c0392b") +
  labs(title = "Marital Satisfaction (+)", x = "Marital Satisfaction", y = "Life Satisfaction") +
  theme_minimal(base_size = 12)

p2 <- ggplot(df, aes(x = partner_conflict, y = life_satisfaction_2)) +
  geom_point(color = "#8e44ad", alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "#c0392b") +
  labs(title = "Partner Conflict (\u2212)", x = "Partner Conflict", y = "Life Satisfaction") +
  theme_minimal(base_size = 12)

library(patchwork)
p1 + p2
```

::: {.callout-note title="Interpretation Summary" icon="false"}
- **Marital satisfaction** shows a **positive** slope ($\beta_1 = `r round(coef(model_1)[2], 2)`$): higher relationship satisfaction is associated with higher life satisfaction.
- **Partner conflict** shows a **negative** slope ($\beta_1 = `r round(coef(model_2)[2], 2)`$): higher conflict is associated with lower life satisfaction.
- Both models follow the same underlying simple linear regression structure from @sec-regression-overview — only the **sign and magnitude of $\beta_1$** differ, illustrating how the same modeling framework captures relationships in opposite directions.
:::

::: {.callout-important title="Looking Ahead: Moderation" collapse="true"}
These two bivariate relationships set up a natural extension: what if the strength of the marital-satisfaction → life-satisfaction relationship (or the conflict → life-satisfaction relationship) depends on a third variable — say, years married, or social support? That question requires adding an interaction term, covered in @sec-moderator-effects.
:::

### The Regression Model

General form of multiple linear regression:

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_p X_p + \varepsilon$$

Where:

- $Y$ = Dependent variable
- $X_1, X_2, \dots, X_p$ = Independent variables
- $\beta_0$ = Intercept (constant)
- $\beta_1, \beta_2, \dots, \beta_p$ = Regression coefficients
- $\varepsilon$ = Error term (random error)

::: {.callout-note title="Interpreting Coefficients" icon="false"}
Coefficients ($\beta$) show the expected change in $Y$ for a one-unit change in $X$, **holding other variables constant**.
:::

------------------------------------------------------------------------

### Goodness of Fit

::::::: {layout-ncol="2"}
::: {.callout-tip title="R² — Coefficient of Determination"}
Proportion of variance in Y explained by the model.
:::

::: {.callout-tip title="Adjusted R²"}
Adjusted for the number of predictors in the model.
:::

::: {.callout-tip title="RMSE — Root Mean Square Error"}
Measures the average prediction error.
:::

::: {.callout-tip title="SEE — Standard Error of Estimate"}
Indicates the typical distance between observed and predicted values.
:::
:::::::

------------------------------------------------------------------------

### Assumptions (Linear Regression)

- **Linearity:** The relationship between X and Y is linear.
- **Independence:** Observations are independent.
- **Homoscedasticity:** Constant variance of errors.
- **Normality:** Errors are normally distributed.
- **No Multicollinearity:** Predictors are not highly correlated with each other.

------------------------------------------------------------------------



### Limitations

::: {.callout-warning title="Use With Caution" icon="false"}
- Sensitive to **outliers**
- Does **not imply causation**
- Model may perform poorly if **assumptions are violated**
- **Overfitting** reduces predictive ability
- **Extrapolation** beyond the data range can be risky
:::

------------------------------------------------------------------------
