---
title: "Chapter 4: Complex Between Group Analyses"
format: revealjs
---

```{=html}
<style>
.reveal{font-size:28}
</style>
```

------------------------------------------------------------------------

### Moving from simple to complex questions

- Determine when to use:

  - One-Way ANOVA
  - Factorial ANOVA
  - Repeated Measures ANOVA
  - Mixed ANOVA
  - ANCOVA
  - MANOVA

- Understand how ANOVA connects to regression and mixed models.

  | 

  ### The Limitation of t-Tests \|

  Suppose we want to compare: 
  - Freshmen 
  - Sophomores 
  - Juniors 
  - Seniors 
  
  Can we use a t-test? NO! 
  
  Also, conducting multiple t-tests creates a problem: 
    - Type I error inflation 
    - The probability of making a false positive conclusion increases with each additional test

### Enter ANOVA

ANOVA stands for:

-**Analysis of Variance**

-Developed by **Sir Ronald Fisher**

Core question:

> Are the observed group differences larger than would be expected due to random variation?

Instead of comparing two means directly, ANOVA compares:

- Between-group variability

to

- Within-group variability

------------------------------------------------------------------------

### The Logic of ANOVA

If groups truly differ:

- Group means should be far apart

and

- Individuals within groups should be relatively similar

ANOVA evaluates:

$$F =
\frac{
\text{Between-Group Variance}
}{
\text{Within-Group Variance}
}$$

------------------------------------------------------------------------

# Understanding the F Ratio

Small F:

$$\text{Between} \approx \text{Within}$$

Little evidence of group differences.

Large F:

$$\text{Between} > \text{Within}$$

Evidence that group means differ.

Think of:

$$F =
\frac{\text{Signal}}{\text{Noise}}$$

just as

$$t =
\frac{\text{Signal}}{\text{Noise}}$$

for a t-test.

------------------------------------------------------------------------

### ANOVA Models that test between groups, within-groups, and mixed models

We will study:

1.  One-Way ANOVA - Outcome compared across multiple groups
2.  Factorial ANOVA - Outcome compared across multiple crossed groups
3.  Repeated Measures ANOVA - Outcome compared across repeated measure or assessment, so this is more within-person than between
4.  Mixed ANOVA - combined within and between
5.  ANCOVA - Comparing group differences after adding covariate
6.  MANOVA - Comparing multiple outcomes at same time
7.  Linear Mixed Models - Regressions that we will do later

Each model extends the previous one.

------------------------------------------------------------------------

### One-Way ANOVA

Does exercise motivation differ across:

- Freshmen
- Sophomores
- Juniors
- Seniors

Independent Variable:

- Class standing

Dependent Variable:

- Exercise motivation

One categorical predictor

One continuous outcome, so mean difference comparison across groups

------------------------------------------------------------------------

### One-Way ANOVA Structure

$$Y = \mu + Group + Error$$

Where:

- Y = outcome
- μ = grand mean
- Group = treatment effect
- Error = residual variation

Question:

> Are some group means different?

------------------------------------------------------------------------

### After One-Way ANOVA

A significant F-test tells us:

> At least one group differs.

It does NOT tell us:

> Which groups differ.

Follow-up analyses include:

- Tukey HSD
- Bonferroni tests
- Planned contrasts

------------------------------------------------------------------------

### Factorial ANOVA

Now suppose we have:

- Gender
- Class Standing

and want to predict:

- Exercise Motivation

We now have two independent variables.

------------------------------------------------------------------------

### Factorial ANOVA Questions

Main Effect of Gender:

> Do males and females differ?

Main Effect of Class Standing:

> Does motivation differ across class levels?

Interaction:

> Does the effect of gender depend on class standing?

------------------------------------------------------------------------

### Interaction Effects

Main effects are often not the most interesting finding.

Interactions answer:

> Does the impact of one variable change across levels of another variable?

Examples:

- Treatment effectiveness differs by gender
- Intervention effects differ by age
- Program effects differ by ethnicity

Interactions are foundational for modern statistical modeling.

------------------------------------------------------------------------

### Repeated Measures ANOVA

Now suppose participants are measured:

- Beginning of semester
- Midterm
- End of semester

Same participants repeatedly measured.

Question:

> Does the outcome change across time?

------------------------------------------------------------------------

### Connection to Paired t-Tests

Paired t-Test:

$$2$$

time points

Repeated Measures ANOVA:

$$3+$$

time points

A paired t-test is simply a special case of repeated measures ANOVA.

------------------------------------------------------------------------

### Mixed ANOVA

Suppose we have:

Groups:

- Treatment
- Control

Measured at:

- Pretest
- Posttest
- Follow-up

Now we have:

Between-Subjects Factor:

- Group

Within-Subjects Factor:

- Time

------------------------------------------------------------------------

# Mixed ANOVA Questions

Main Effect of Time:

> Do scores change over time?

Main Effect of Group:

> Do groups differ overall?

Interaction:

> Do groups change differently over time?

The interaction is usually the primary research question.

------------------------------------------------------------------------

### ANCOVA

Analysis of Covariance

Research Question:

> Do interventions differ after controlling for baseline differences?

Example:

Outcome:

- Posttest depression

Group:

- Treatment vs Control

Covariate:

- Baseline depression

------------------------------------------------------------------------

### ANCOVA Model

ANCOVA combines:

ANOVA +

Regression

Conceptually:

$$Outcome =
Group + Covariate + Error$$

ANCOVA adjusts group means for preexisting differences.

------------------------------------------------------------------------

### MANOVA

MANOVA stands for:

**Multivariate Analysis of Variance**

ANOVA:

- One outcome

MANOVA:

- Multiple outcomes

------------------------------------------------------------------------

### Example MANOVA

Independent Variable:

- Treatment Group

Dependent Variables:

- Depression
- Anxiety
- Stress
- Self-Esteem

Question:

> Do groups differ on the overall profile of outcomes?

------------------------------------------------------------------------

### Why Use MANOVA?

Instead of conducting:

- ANOVA for Depression
- ANOVA for Anxiety
- ANOVA for Stress
- ANOVA for Self-Esteem

MANOVA evaluates them simultaneously.

Advantages:

- Controls Type I error
- Examines combined outcome patterns

------------------------------------------------------------------------

# Linear Mixed Models

Repeated-measures ANOVA has limitations:

- Missing data
- Unequal measurement occasions
- Unequal spacing of observations
- Nested data structures

Mixed models address these issues.

------------------------------------------------------------------------

# Examples of Mixed Models

Students nested within classrooms

Repeated assessments nested within individuals

Patients nested within clinics

Family members nested within families

Longitudinal developmental studies

------------------------------------------------------------------------

# Mixed Model Example

$$Y_{ij}
=
\beta_0
+
\beta_1(Time)
+
u_i
+
e_{ij}$$

Where:

- Fixed effects describe average change
- Random effects describe individual differences

------------------------------------------------------------------------

# Big Picture

t-Test

↓

One-Way ANOVA

↓

Factorial ANOVA

↓

Repeated Measures ANOVA

↓

Mixed ANOVA

↓

ANCOVA

↓

MANOVA

↓

Linear Mixed Models

------------------------------------------------------------------------

# Example 4: One-Way ANOVA

## Research Question

Researchers were interested in whether ethnic-racial identity (ERI) exploration differed across stages of high school.

Specifically, they compared:

- 9th Grade Students
- 10th Grade Students
- 11th Grade Students
- 12th Grade Students

Outcome:

- ERI Exploration Score

Research Question:

> Does ethnic-racial identity exploration differ across grade levels?

------------------------------------------------------------------------

# Why Not Use Multiple t-Tests?

Suppose we compared:

- 9th vs 10th
- 9th vs 11th
- 9th vs 12th
- 10th vs 11th
- 10th vs 12th
- 11th vs 12th

That would require six separate t-tests.

Problems:

- Increased Type I Error
- More complicated interpretation

Instead, we use a single One-Way ANOVA.

------------------------------------------------------------------------

# Independent Variable

## Grade Level

| Group |
|-------|
| 9th   |
| 10th  |
| 11th  |
| 12th  |

This variable is categorical.

------------------------------------------------------------------------

# Dependent Variable

## ERI Exploration

Example items:

- I have spent time trying to learn more about my ethnic group.
- I have participated in activities that help me understand my ethnic background.
- I think a lot about what my ethnicity means to me.

Higher scores indicate greater exploration.

------------------------------------------------------------------------

# Hypotheses

Null Hypothesis:

$$H_0:
\mu_{9th}
=
\mu_{10th}
=
\mu_{11th}
=
\mu_{12th}$$

All population means are equal.

Alternative Hypothesis:

$$H_A:
\text{At least one population mean differs}$$

Notice that ANOVA does not specify which group differs.

------------------------------------------------------------------------

# Sample Data

Suppose the study produced the following sample means.

| Grade |   N | Mean |
|-------|----:|-----:|
| 9th   |  50 | 2.90 |
| 10th  |  50 | 3.12 |
| 11th  |  50 | 3.47 |
| 12th  |  50 | 3.61 |

Visual inspection suggests exploration may increase across grade levels.

The question is:

> Are these differences larger than we would expect from sampling error alone?

------------------------------------------------------------------------

# Logic of ANOVA

ANOVA compares:

## Between-Group Variation

Differences among group means

versus

## Within-Group Variation

Differences among individuals inside each group

The test statistic is

$$F=
\frac{
MS_{Between}
}{
MS_{Within}
}$$

------------------------------------------------------------------------

# Understanding the F Ratio

If group means are very similar:

$$F \approx 1$$

Little evidence of group differences.

If group means differ substantially:

$$F > 1$$

Evidence of group differences.

Larger F values provide stronger evidence against the null hypothesis.

------------------------------------------------------------------------

# Simulating Data

```{r}
set.seed(123)

n <- 50

dat <- data.frame(
  grade = factor(rep(
    c("9th","10th","11th","12th"),
    each = n
  )),
  exploration = c(
    rnorm(n, 2.90, .50),
    rnorm(n, 3.12, .50),
    rnorm(n, 3.47, .50),
    rnorm(n, 3.61, .50)
  )
)

head(dat)
```

------------------------------------------------------------------------

# Descriptive Statistics

```{r}
library(dplyr)

dat %>%
  group_by(grade) %>%
  summarise(
    N = n(),
    Mean = mean(exploration),
    SD = sd(exploration)
  )
```

------------------------------------------------------------------------

# Conducting the One-Way ANOVA

```{r}
fit <- aov(
  exploration ~ grade,
  data = dat
)

summary(fit)
```

------------------------------------------------------------------------

# Example Output

``` text
             Df Sum Sq Mean Sq F value Pr(>F)
grade         3  18.72   6.24   21.84 <.001
Residuals   196  56.02   0.29
```

Interpretation:

$$F(3,196)=21.84,\quad p<.001$$

The omnibus test indicates significant differences among grade levels.

------------------------------------------------------------------------

# What Does the Omnibus Test Tell Us?

The ANOVA tells us:

✅ Some mean differs

The ANOVA does NOT tell us:

❌ Which means differ

To answer that question we conduct post hoc comparisons.

------------------------------------------------------------------------

# Tukey Post Hoc Tests

```{r}
TukeyHSD(fit)
```

------------------------------------------------------------------------

# Example Interpretation

Suppose the results indicate:

- 11th \> 9th
- 12th \> 9th
- 12th \> 10th

Interpretation:

Students in the upper grades reported significantly greater ethnic-racial identity exploration than students in the lower grades.

------------------------------------------------------------------------

# Effect Size

One common effect size is Eta-Squared:

$$\eta^2
=
\frac{
SS_{Between}
}{
SS_{Total}
}$$

In R:

```{r}
library(effectsize)

eta_squared(fit)
```

------------------------------------------------------------------------

# APA Style Write-Up

A one-way analysis of variance examined differences in ethnic-racial identity exploration across grade levels. Results indicated a significant effect of grade level on exploration, F(3, 196) = 21.84, p \< .001. Follow-up Tukey comparisons suggested that students in the upper grades reported greater levels of ethnic-racial identity exploration than students in the lower grades.

------------------------------------------------------------------------

# Key Takeaways

- One-Way ANOVA compares three or more means.
- The F-test compares between-group variability to within-group variability.
- A significant omnibus test indicates that at least one mean differs.
- Post hoc tests identify where differences occur.
- Effect sizes quantify the magnitude of group differences.

# 
