---
title: "Data Cleaning and Using R Slides"
format: pptx
---

## Using R to Read Data, Recode Variables, and Prepare Data for Analysis

Usually to get started with some sort of data analysis we will follow these steps:

1.  Read data into R
2.  View and inspect data
3.  Clean any data issues or problems
4.  Recode where necessary
5.  Create summary variables if necessary
6.  Standardizing variables
7.  Decisions regarding missing data

## Reading data into R

Base R has a number of built-in functions for reading in text and ASCII data

- Some key terms: Data are *delimited* or separated in text files by things like *tabs*, *commas*, or *spaces*
- How the data are delimited influences which function should be used
- For example, if I want to read in a comma separated file (.csv) that I have saved using Excel or notepad
  - data_file \<- read.csv("stored_data.csv")

## Example using R Code

```{r}
#| echo: true 
#| eval: false
Hmong_df <- read.csv("HmongData.csv") # read.csv
head(Hmong_df)                        # Show the first few lines
str(Hmong_df)       # structure
# names(Hmong_df)     # variable names - too long for now


```

## Quick Cheatsheet

- read.csv(): CSV files

- read.delim(): Tab-delimited files

- read.table(): General text files

- read.fwf(): Fixed-width files load(): .RData files

- readRDS(): .rds files file.choose(): Browse for file

- note the file.choose() - this allows for opening a folder, and then clicking in the file

## File.choose option

```{r}
#| echo: true 
#| eval: false
Hmong_df<-read.csv(file.choose())
# Opens a file browser so you can point and click # to select a data file
\
```

## Full workup for the Hmong .csv file

Note that the HmongData.csv has missing data values coded as -99

- Need to communicate that to R
- There are also options regarding whether the first row contains variable names
  - header = TRUE or header = FALSE
  - Since header = T is the default, I omit it in my code

```{r}
#| echo: true 
#| eval: false

Hmong_df <- read.csv("HmongData.csv",na.strings=-99)

head(Hmong_df)


```

## Using Packages to open data sets

*readr*

- readr is part of the tidyverse, so loading tidyverse also loads other packages like ggplot2

- library(readr) has the the following functions

- read_csv

- read_tsv

- read_delim

- These produce tibbles, which are a "modern version" of data frames designed to work with tidyverse packages

  - They also, when viewed, indicate the type of each variable in the data (integer, numeric, character, logical, date, and more)

## readr workflow

```{r}
#| echo: true 
#| eval: false
library(readr)
Hmong_df2 <- read_csv("HmongData.csv",na="-99")
Hmong_df2

table(Hmong_df2$BornUS)

```

## Reading Excel files

```{r}
#| echo: true 
#| eval: false
library(readxl)


Hmong_df3 <- read_excel("HmongData.xlsx", na = "-99")

head(Hmong_df3)

```

## Haven package to read in SPSS, SAS, Stata files

library(haven)

- df \<- read_sav("mydata.sav")

- df \<- read_dta("mydata.dta")

- df \<- read_sas("mydata.sas7bdat")

- Haven automatically imports missing values and metadata stored in SPSS, SAS, and Stata files

  - Things like variable labels, value labels

## Haven for SPSS file

```{r}
#| echo: true 
#| eval: false
library(haven)
Hmong_df4 <- read_sav("HmongData.sav")
head(Hmong_df4)

```

## Checking data

1.  Outliers
2.  Miscoded or data entry errors
3.  Missing data codes are correct

- psych package
  - describe(Hmong_df)
  - gives overall means, skew, range, minimum/maximum, etc
  - Use the output to identify out-of-range values (some odd values, -99 for example)
  - Outliers, etc.

## Recoding variables

- Reverse scoring any items
  - The valence of an item *positive or negative* to keep variables in a direction that is theoretically relevant
  - Sometimes scoring is done in a way that a greater number actually is *less* of a variable
- Coding any binary variables
  - Recommended that these are 0/1, particularly in regressions
  - R has no problem using factor variable, Regression models will automatically choose a reference category unless one is specified.
- Nominal variables
  - Dummy coding necessary in regressions

## Summary scores

Often in HHS research we use composites, sum scores to represent *psychological* constructs

- Anxiety, depression, exercise motivation, racial identity, anything that we think captures some phenomenon of interest
- Often these are **latent constructs** represented by *indicators* or items
- The GAD-7 *Generalized Anxiety Disorder* scale includes 7 indicators
  - Logic is that someone's underlying anxiety would *cause* them to indicate more frequent experiences of **feeling nervous, anxious, or on edge**, **having trouble relaxing**, etc
- Typically researchers administer to participants and then do some summation of the items
  - Most simple is to add items together for a sum score (naive or coarse measurement)
  - Sometimes researchers average the items to stay on a "scale", in this case 0-3
  - Can use other approaches using factor analysis and weighting items
- Can use BaseR functions or packages like psych to create these

## Standardizing

There are some advantages to standardizing numeric, continuous variables

- Can have two versions of the same variable, standardized (SD units) and unstandardized (raw units)

- When we run regressions and want standardized output, some advantages of just using standardized outcomes

- in R, can use the scale() function

  - data_df$anxiey_z <- scale(data_df$anxiety, center = T, scale = T)
    - To center can use scale = F
    - Center is subtracting sample mean scores from individual scores
      - Results in a mean of zero, but scale stays the same
      - Standardization results in mean of zero, but SD/Var is 1
